---
title: "Custom plots from bicorX results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom plots from bicorX results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why this vignette exists

Earlier versions of bicorX exported around 40 plotting functions - one
for nearly every combination of method (BMHT/BMKC/ROS-DET) and data
modality (bulk, single-cell, spatial, microbiome). Most of them were
thin wrappers around the same few ideas (a volcano plot, a network
diagram, a heatmap, an embedding overlay) applied to slightly different
inputs, and each one was a function signature, a help page, and a
maintenance burden that almost nobody used.

As of this release, the package ships eight core plotting functions,
covering the visualizations most people reach for first:

| Function | Method | What it shows |
|---|---|---|
| `plot_rewiring_volcano()` | BMHT | DC score vs. -log10(FDR) |
| `plot_top_gene_bar()` | BMHT | Top rewired genes, ranked |
| `plot_module_heatmap()` | BMKC | Before/after correlation heatmap for a module |
| `plot_module_eigengene()` | BMKC | Module eigengene across conditions |
| `plot_umap_eigengene()` | BMKC | Module eigengene projected onto a UMAP (needs Seurat) |
| `plot_differential_network()` | BMHT & BMKC | Rewired-edge network diagram |
| `plot_rosdet_quadrant()` | ROS-DET | Correlation in condition A vs. B, per pair |
| `plot_edge_volcano()` | ROS-DET | Distance score vs. -log10(FDR), per pair |

Everything else - taxa/microbiome breakdowns, causal-driver overlays,
hive plots, 3D networks, spatial coordinate overlays, chord diagrams,
raincloud plots, and so on - is one `ggplot2`/`igraph`/`Seurat` call away
from the result object itself. This vignette shows the pattern, using
the same data every result object already carries.

## What's in a result object

Every `run_bmht()`/`run_bmkc()`/`run_rosdet()` call returns a list with a
`results` data.frame (the per-gene or per-pair statistics) and a `data`
list (the input expression matrix, condition vector, and species vector,
if any) - so you always have everything needed to build a custom view.

```{r, eval = FALSE}
library(bicorX)

res_bmht <- run_bmht(expr_matrix, condition, n_permutations = 1000)

names(res_bmht)
#> [1] "results" "n_permutations" "half_threshold" "significance_level"
#> [5] "species_mode" "transform" "data"

head(res_bmht$results)
#>       Gene DC_Score    P_value        FDR
#> 1 Gene_007 2.549791 0.00990099 0.01094032
#> ...

str(res_bmht$data, max.level = 1)
#> List of 3
#>  $ expr_matrix: num [1:200, 1:24] ...
#>  $ condition  : Factor w/ 2 levels "A","B": 1 1 1 ...
#>  $ species    : NULL
```

ROS-DET results have the same shape, but `results` is a per-pair table
(`Gene1`, `Gene2`, `Distance_Score`, `P_value`, `FDR`, `r1`, `r2`,
`c_weight`) and there's an additional `n_tested` field. BMKC results
replace `results` with a `modules` list (one character vector of gene
names per module) and an `n_modules` count.

## Example 1: a taxa/microbiome breakdown (was `plot_taxa_rewiring`)

If you're working with compositional data and have a taxonomy table on
the side, join it to `results` by gene/feature name and facet or color
by rank - there's no need for a dedicated function:

```{r, eval = FALSE}
library(ggplot2)

# taxa_df: a data.frame with columns Gene, Phylum (or your rank of choice)
merged <- merge(res_bmht$results, taxa_df, by = "Gene")

ggplot(merged, aes(x = reorder(Gene, DC_Score), y = DC_Score, fill = Phylum)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ Phylum, scales = "free_y") +
  labs(x = NULL, y = "DC Score", title = "Rewiring by taxonomic group") +
  theme_bicor()
```

## Example 2: a causal-driver / ego network (was `plot_causal_network`, `plot_ego_switching`)

`plot_differential_network()` already builds an `igraph` object
internally; the same recipe - recompute condition-specific correlation
matrices, threshold the difference, build a graph - generalizes to any
custom layout, coloring, or subsetting you want. To focus on one gene's
neighborhood (an "ego network"):

```{r, eval = FALSE}
library(igraph)

genes <- head(res_bmht$results$Gene, 30)
expr_sub <- res_bmht$data$expr_matrix[genes, , drop = FALSE]
cond <- res_bmht$data$condition

cor1 <- cpp_calc_observed_bicor(expr_sub[, cond == levels(cond)[1]])
cor2 <- cpp_calc_observed_bicor(expr_sub[, cond == levels(cond)[2]])
rownames(cor1) <- colnames(cor1) <- rownames(cor2) <- colnames(cor2) <- genes

adj_diff <- cor2 - cor1
adj_diff[abs(adj_diff) < 0.3] <- 0

g <- graph_from_adjacency_matrix(adj_diff, mode = "undirected", weighted = TRUE, diag = FALSE)

# Ego network: just the target gene and its immediate rewired neighbors
target <- genes[1]
ego_nodes <- c(target, neighbors(g, target)$name)
g_ego <- induced_subgraph(g, ego_nodes)

plot(g_ego, vertex.color = ifelse(V(g_ego)$name == target, "orange", "steelblue"),
     edge.color = ifelse(E(g_ego)$weight > 0, "red3", "steelblue4"),
     main = paste("Rewired neighbors of", target))
```

## Example 3: a UMAP overlay for a different statistic (was `plot_umap_disruption_landscape`, `plot_umap_dual_cliques`, `plot_umap_edge_activity`)

`plot_umap_eigengene()` covers the single most common case (module
eigengene on a UMAP). For anything else - a per-cell disruption score, a
dual-module comparison, or an edge-activity score - the pattern is the
same: compute a per-cell score, add it as a Seurat metadata column or
module score, and let `Seurat::FeaturePlot()` do the rest.

```{r, eval = FALSE}
library(Seurat)

top_genes <- head(res_bmht$results$Gene, 20)
valid_genes <- intersect(top_genes, rownames(sc_obj))

sc_obj <- AddModuleScore(sc_obj, features = list(valid_genes), name = "Disruption")
FeaturePlot(sc_obj, features = "Disruption1", reduction = "umap") +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0)
```

## Example 4: a spatial overlay (was `plot_spatial_coords`, `plot_spatial_density`, `plot_spatial_tissue`, `plot_spatial_edge`, `plot_network_proximity`)

If your data has spatial coordinates (a `Seurat` spatial object, or a
plain data.frame of x/y positions), the same "compute a per-feature
score, then map it onto coordinates" pattern applies:

```{r, eval = FALSE}
genes <- res_bmkc$modules[[1]]
module_expr <- colMeans(res_bmkc$data$expr_matrix[genes, , drop = FALSE])

spatial_df <- data.frame(x = spatial_coords[, 1], y = spatial_coords[, 2],
                          score = module_expr[rownames(spatial_coords)])

ggplot(spatial_df, aes(x, y, color = score)) +
  geom_point(size = 2) +
  scale_color_viridis_c() +
  coord_fixed() +
  labs(title = "Module activity across spatial coordinates") +
  theme_bicor()
```

## Example 5: comparing results across methods (was `plot_method_overlap`)

```{r, eval = FALSE}
library(UpSetR)

sig_bmht <- res_bmht$results$Gene[res_bmht$results$FDR < 0.05]
sig_rosdet <- unique(c(res_rosdet$results$Gene1, res_rosdet$results$Gene2)
                      [res_rosdet$results$FDR < 0.05])
sig_bmkc <- unique(unlist(res_bmkc$modules))

upset(fromList(list(BMHT = sig_bmht, `ROS-DET` = sig_rosdet, BMKC = sig_bmkc)),
      order.by = "freq")
```

## A note on `theme_bicor()` and `bicor_palette()`

Both are still exported and designed to be dropped into any custom
`ggplot2` call, so your own plots match the look of the built-in ones:

```{r, eval = FALSE}
ggplot(res_bmht$results, aes(DC_Score, -log10(FDR))) +
  geom_point(color = bicor_palette()["Condition1"]) +
  theme_bicor()
```
