---
title: "Getting started: data types, transforms, and choosing a significance method"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: data types, transforms, and choosing a significance method}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette is the reference for "what kind of data can I put into this
package, and how do I get a result I can trust from it." For a worked
cross-layer example, see `vignette("multiomics-workflow")`; for rebuilding
removed plot types, see `vignette("custom-plots")`.

## The three methods, briefly

- **`run_bmht()`** - per-gene score: does this gene's overall correlation
  structure change between conditions? Good for a ranked gene list.
- **`run_rosdet()`** - per-pair score: does this specific pair of genes
  flip from positively to negatively correlated? Good for finding
  individual "switching" relationships.
- **`run_bmkc()`** - module-level: which groups of genes form a
  differentially co-expressed community? Good for a higher-level summary
  than a gene or pair list.

All three take the same basic shape of input: an expression matrix
(genes/features as rows, samples as columns) and a two-level condition
factor.

## Input types

You don't have to hand-build a plain matrix. `run_bicordcea()` (and the
three method-specific wrappers, which call it internally) dispatches on
several common object types:

```{r}
library(bicorX)

# Plain matrix or data.frame
run_bmht(expr_matrix, condition)

# Bioconductor SummarizedExperiment - condition pulled from colData
run_bmht(se_object, condition = "treatment_group")

# Seurat object
run_bmht(seurat_object, condition = "cell_type")

# MultiAssayExperiment (for cross-omics designs - see the
# multiomics-workflow vignette for the fuller picture)
run_bmht(mae_object, condition = "subtype")
```

### Single-cell and spatial data: aggregate first

Bicor-based differential co-expression is a bulk-style method - it
expects one row per sample, not one row per cell. For single-cell or
spatial data, aggregate to pseudobulk profiles first:

```{r}
pb <- make_pseudobulk(sce_object, sample_col = "patient_id",
                       condition_col = "treatment_group", min_cells = 10)
run_bmht(pb$expr_matrix, factor(pb$condition))
```

`sample_col` should identify independent biological replicates (patients,
mice, wells) - not cell clusters or cell types. Aggregating within a
cluster/cell-type instead of within a real replicate loses the whole
point of pseudobulking (avoiding pseudo-replication across cells from the
same sample), and gives you no real statistical replication if that
cluster only exists in one specimen. If your data genuinely has no
multi-replicate structure (e.g., a single demo dataset with only one
specimen), you can still run the pipeline for a code-path check, but
don't treat the result as a validated finding - see
`dev/real_data_validation/singlecell_pathway_check.R` for exactly this
distinction, worked through in full.

## Transforms: when do you need one?

| `transform =` | Use for | Requires |
|---|---|---|
| `"none"` (default) | Any continuous data - microarray, log-ratios, z-scores, already-normalized RNA-seq | Nothing extra |
| `"log1p"` | Raw, non-negative count data without a compositional structure | Non-negative values |
| `"clr"` | Compositional data where the *total* (library size, total taxa count) isn't itself meaningful - microbiome, ATAC peak counts | Non-negative values |
| `"rclr"` | Same as `"clr"` but robust to the zero-inflation compositional data often has | Non-negative values |

**Why this matters, concretely**: raw counts confound total sequencing
depth with true relative abundance. Two features that both happen to
scale with a sample's own depth will show spurious correlation with each
other regardless of real biology - and this can differ between
conditions even without an average depth difference between them, purely
from within-condition variance in per-sample depth. `clr`/`rclr` remove
this by working in per-sample log-ratio space instead of raw counts. This
isn't a theoretical concern: on real published lung microbiome data, a
naive analysis produced a spurious, highly "significant" finding
(FDR = 5.5e-06) between two taxa that was fully explained by exactly this
per-sample depth artifact, and CLR correction removed it. See
`dev/real_data_validation/lungdata_compositional_check.R` for the full,
reproducible investigation, including a second claim that looked
promising but was checked with a formal test and correctly rejected -
worth reading if you want to see what "validated" actually means in this
package's own development process, not just take the claim on faith.

### Multi-layer data

If you have two feature sets on the same samples (RNA + ATAC, host +
microbe, two different omics platforms), CLR/rCLR is computed
*separately within each layer* rather than globally, using the
`layer`/`species` argument - see `vignette("multiomics-workflow")` for
the full mechanism and a real (TCGA RNA+miRNA) worked example, not just
a synthetic one.

## Choosing a significance method (ROS-DET only, for now)

`run_rosdet()` has two ways to get a p-value:

```{r}
# Default: exact permutation test, no distributional assumptions,
# resolution capped by n_permutations
run_rosdet(expr_matrix, condition, n_permutations = 5000)

# Analytical (Kayano et al. 2011's own "ECOR" test): continuous
# resolution at a small, fixed cost per pair - no permutation needed
run_rosdet(expr_matrix, condition, significance_method = "analytical")
```

**Use `"analytical"` when you have at least ~30-50 samples per
condition** - below that it becomes progressively (though not wildly)
anti-conservative, and a warning fires below N=30. Above that threshold,
it's both faster and better-resolved than permutation testing, and this
isn't a marginal difference: on a real published ground-truth benchmark,
switching to `"analytical"` took ROS-DET's ranking quality (AUPRC) from
0.21 - roughly 3.5x worse than two established competitors (DGCA, dcanr,
both around 0.73) - to 0.69, roughly on par with them. See
`dev/benchmarks/03_sim102_ground_truth.R` for the full, reproducible
comparison.

**Use `"permutation"`** (the default, unchanged from this package's
original behavior) when you have small samples, or want zero
distributional assumptions regardless of sample size. It's the only
option currently available for `run_bmht()` and `run_bmkc()` - their
source papers don't specify an analytical alternative the way ROS-DET's
does, so this is a genuine, not-yet-closed asymmetry between the three
methods, not an oversight in this vignette.

## Performance notes

- **BLAS matters more than you'd expect.** This package's cost is
  dominated by a correlation-matrix computation that a plain reference
  BLAS handles slowly. Installing OpenBLAS gives a measured, real
  ~1.3-1.4x speedup at zero code changes - see `DESCRIPTION`'s
  `SystemRequirements` field.
- **Multi-core scales automatically** via the `workers` argument
  (OpenMP) - set it to your available core count for permutation-heavy
  analyses. This isn't just a suggestion: the two competing tools this
  package has been benchmarked against (DGCA, dcanr) have no equivalent
  parallelism in their own permutation code, so this is a real,
  structural advantage on multi-core hardware, not a marginal tuning
  knob.
- **Memory no longer scales with `n_permutations`** for BMHT/ROS-DET -
  large permutation counts (tens of thousands) are practical without
  the memory footprint growing proportionally.

None of this closes a statistical resolution gap on its own - it makes
running a genuinely adequate number of permutations (or, for ROS-DET,
using the analytical method) practical rather than prohibitively slow.

## Where the evidence for all of this actually lives

Every specific number in this vignette traces back to a reproducible
script, not just a claim:

- `dev/benchmarks/` - the original outlier-robustness and ground-truth
  ranking-quality benchmarks (Phase 5, Phase A).
- `dev/real_data_validation/` - real (not synthetic) validations: lung
  microbiome (compositional correction), TCGA cancer data (multi-omics),
  and single-cell (pathway check only - explicitly not the same
  statistical footing as the other two, and the script says so directly).
- `NEWS.md` - the full chronological record, including negative results
  (four rejected attempts at a different fix for ROS-DET's permutation
  resolution, and one enrichment claim that looked real but wasn't)
  reported honestly rather than discarded.
