---
title: "Multi-omics workflow: combining RNA and ATAC (or host + microbe) data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-omics workflow: combining RNA and ATAC (or host + microbe) data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What this is for

bicorX's most distinctive capability isn't any one of BMHT, BMKC, or
ROS-DET individually - it's that all three can be pointed at data that
combines more than one omics layer (RNA + ATAC, host + microbe, or any
two feature sets measured on the same samples) and asked specifically
about *cross-layer* relationships: does this gene's expression rewire
alongside this regulatory element's accessibility, and only in one
condition?

This vignette walks through that workflow end to end using a small
built-in synthetic dataset, then explains the mechanism (per-layer CLR
normalization) that makes it work correctly rather than just concatenating
two unrelated matrices and hoping for the best.

## The built-in example dataset

`inst/extdata/multiomics_dataset.RData` contains a synthetic RNA layer (30
genes) and ATAC layer (20 peaks) across 24 samples in two conditions, with
a deliberately injected signal: 5 genes and 5 peaks share a common latent
factor in condition A only - as if a set of regulatory elements becomes
active and drives a set of genes, specifically under one condition.

```{r, eval = FALSE}
library(bicorX)

load(system.file("extdata", "multiomics_dataset.RData", package = "bicorX"))
dim(multiomics_rna)   # 30 genes x 24 samples
dim(multiomics_atac)  # 20 peaks x 24 samples
table(multiomics_condition)
```

## Running the analysis

`run_multiomic()` is the entry point: it takes two matrices sharing the
same samples, tags each feature with which layer it came from, and runs
any of the three methods restricted to cross-layer relationships by
default.

```{r, eval = FALSE}
res <- run_multiomic(
  multiomics_rna, multiomics_atac, multiomics_condition,
  method = "rosdet",       # also supports "bmht" and "bmkc"
  n_permutations = 2000,   # candidate-pair counts here can be in the
                           # hundreds, so this needs to be well above the
                           # ROS-DET default to survive FDR correction -
                           # see the note on permutation counts below
  min_delta = 0.3
)

head(res$results[order(res$results$FDR), ])
```

Feature names are prefixed with the layer they came from
(`Layer_A_Gene_01`, `Layer_B_Peak_01` by default, or your own
`layer_A_name`/`layer_B_name`), so it's always unambiguous which pairs are
cross-layer:

```{r, eval = FALSE}
sig <- res$results[res$results$FDR < 0.05, ]
sig[, c("Gene1", "Gene2", "Distance_Score", "FDR")]
```

## `pair_type_filter`: inter, intra, or all

By default `run_multiomic()` sets `pair_type_filter = "inter"`, keeping
only cross-layer pairs (RNA-vs-ATAC), which is almost always what you want
for a multi-omics question - within-layer pairs (gene-vs-gene,
peak-vs-peak) are just the ordinary single-omics analysis and don't need
two layers to detect. Set `pair_type_filter = "all"` to keep everything,
or `"intra"` to keep only within-layer pairs. This filter only applies to
ROS-DET (a pairwise method); it's ignored - with a warning - if supplied
for `method = "bmht"` or `"bmkc"`, which don't produce a pairwise table.

## `layer` vs. `species`: same mechanism, clearer name

Under the hood, `run_multiomic()` is a thin wrapper around
`run_bicordcea()` that builds a `layer` vector (one entry per combined
row) and passes it through. If you're combining matrices yourself instead
of using `run_multiomic()`, `layer` works directly on any of the standard
entry points:

```{r, eval = FALSE}
combined <- rbind(multiomics_rna, multiomics_atac)
layer_tags <- c(rep("RNA", nrow(multiomics_rna)), rep("ATAC", nrow(multiomics_atac)))

run_bmht(combined, multiomics_condition, layer = layer_tags, n_permutations = 1000)
```

`layer` is a synonym for `species` (the original argument name, oriented
toward microbiome use) - both populate the same internal grouping
mechanism. Use whichever name matches what you're actually doing; if you
supply both, `layer` wins.

## Why per-layer, not global, normalization

If you request a compositional transform (`transform = "clr"`), it's
computed *separately within each layer*, not globally across the combined
matrix. This matters because RNA counts and ATAC peak accessibility (or
host and microbial abundances) live on different scales and have
different sparsity patterns - a single global CLR would let whichever
layer has larger or more variable values dominate the normalization and
distort the other layer's compositions. Splitting by `layer`/`species`
keeps each layer's normalization self-contained. See
`?bicorX:::.compositional_transform` for the implementation details.

## A note on permutation counts for cross-layer analyses

Restricting to inter-layer pairs still often leaves hundreds of candidate
pairs to test, and BH-FDR correction across that many tests needs smaller
p-values than a handful of permutations can produce - the smallest
possible p-value at `n_permutations = 100` is `1/101 ≈ 0.0099`, which
usually can't survive correction across 200+ tests. In practice, plan for
`n_permutations` in the thousands for a full multi-omics ROS-DET run;
1,000-5,000 depending on how many candidate pairs survive Gate 1
(`min_delta`).

If you have at least ~30-50 samples per condition, `run_multiomic(...,
significance_method = "analytical")` (passed through to ROS-DET via
`...`) sidesteps this permutation-count tradeoff entirely - it computes
an exact p-value per pair directly rather than by resampling, at the
cost of the small-sample calibration caveats documented on
`?run_rosdet`. This is the significance method used in the real-data
example below.

## A real example: TCGA Adrenocortical Carcinoma (RNA + miRNA)

The workflow above uses a small synthetic dataset for clarity. For a
real, externally-published, adequately-powered example instead, see
`dev/real_data_validation/acc_multiomics_check.R` in the package source
- it runs the exact same `run_multiomic()` workflow on real TCGA
Adrenocortical Carcinoma data (RNA-seq + miRNA-seq, 78 patients,
comparing the published C1A/C1B molecular subtypes), bundled directly
with the Bioconductor `MultiAssayExperiment` package as `miniACC` (no
extra download required). It finds a striking, statistically-confirmed
pattern: a single microRNA (`hsa-mir-137`) accounts for 6 of the 10
genome-wide-significant cross-layer pairs - a massive, formally-tested
concentration (Fisher's exact test p = 5.0e-13 against the correct
background rate), not an eyeballed coincidence. The script's own header
comments are explicit about what's independently verified computationally
within the analysis versus what would need a literature search to confirm
(the specific biological role of that microRNA in this cancer type) -
worth reading in full before citing the result, since it draws that line
carefully rather than overclaiming.
